Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-14 paper/ms.tex

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are popular stochastic
models for time-series analysis.
For GP modeling, a functional form is chosen to describe the autocovariance
of the data and the parameters of this function are fit for or marginalized.
In the astrophysical literature, GPs have been used to model stochastic
variability in light curves of stars \citep{Brewer:2009}, active galactic nuclei \citep{Kelly:2014},
and X-ray binaries (CITE).
They have also been used as models for the cosmic microwave background (CITE),
correlated instrumental noise (CITE), spectroscopic calibration (CITE) and
residuals caused by model inconsistencies (CITE + better words).
While these models are widely applicable, their use has been limited, in
practice, by the computational cost and scaling.
In general, the cost of computing a GP likelihood scales as the third power of
the number of data points $\mathcal{O}(N^3)$ and in the current era of large
time-domain surveys~--~with $\sim10^{4-9}$ targets with $\sim10^{3-5}$
observations each~---~this cost is prohibitive.

>>>>

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are popular stochastic
models for time-series analysis.
For GP modeling, a functional form is chosen to describe the autocovariance
of the data and the parameters of this function are fit for or marginalized.
In the astrophysical literature, GPs have been used to model stochastic
variability in light curves of stars \citep{Brewer:2009}, active galactic nuclei \citep{Kelly:2014},
and the logarithmic flux of X-ray binaries \citep{Uttley:2005}.
They have also been used as models for the cosmic microwave background \citep{Bond:1987,Bond:1999},
correlated instrumental noise \citep{Gibson:2012}, spectroscopic calibration \cite{Czekala:2017,Evans:2015} and
residuals caused by model inconsistencies (CITE + better words).
While these models are widely applicable, their use has been limited, in
practice, by the computational cost and scaling.
In general, the cost of computing a GP likelihood scales as the third power of
the number of data points $\mathcal{O}(N^3)$ and in the current era of large
time-domain surveys~--~with $\sim10^{4-9}$ targets with $\sim10^{3-5}$
observations each~---~this cost is prohibitive.
